---
title: "ST2195 Coursework — Part 1: Random Walk Metropolis Algorithm"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    keep_md: false
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
# Global chunk options — apply to every code chunk unless overridden
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 5,
  dev       = "png"          
)

set.seed(42)  # global random seed for reproducibility
```

---

# Introduction

This report implements and analyses the **Random Walk Metropolis** algorithm
(a specific instance of the Metropolis–Hastings MCMC algorithm) applied to
the **Laplace distribution**:

$$f(x) = \frac{1}{2}\,e^{-|x|}, \qquad x \in \mathbb{R}$$

The Laplace distribution is symmetric around 0 with a sharp peak (heavier
tails than the Normal).  Its true mean is **0** and its true standard
deviation is $\sqrt{2} \approx 1.4142$.

---

# Part 1(a): Sampling from the Laplace Distribution

## Algorithm Description

The **Random Walk Metropolis** algorithm generates a Markov chain
$x_0, x_1, \ldots, x_N$ whose stationary distribution is $f(x)$.

**Step 1.** Choose initial value $x_0$, number of iterations $N$, and
proposal standard deviation $s > 0$.

**Step 2.** For $i = 1, \ldots, N$:

1. Draw a **proposal** $x^* \sim \mathcal{N}(x_{i-1},\, s^2)$.
2. Compute the **log-acceptance ratio** (numerically safer than the raw ratio):
$$\log r(x^*, x_{i-1}) = \log f(x^*) - \log f(x_{i-1}) = -|x^*| + |x_{i-1}|$$
3. Draw $u \sim \text{Uniform}(0,1)$.
4. If $\log u < \log r$, **accept**: set $x_i = x^*$.  
   Otherwise, **reject**: set $x_i = x_{i-1}$.

The full chain has $N + 1$ elements $(x_0, x_1, \ldots, x_N)$.

## Implementation

```{r metropolis_rw}
# -------------------------------------------------------------------
# Log of the target density  log f(x) = -log(2) - |x|
# We only need differences, so the constant -log(2) cancels.
# -------------------------------------------------------------------
log_target <- function(x) -abs(x)

# -------------------------------------------------------------------
# metropolis_rw()
#   N  : number of MCMC steps
#   s  : proposal standard deviation
#   x0 : starting value
#
# Returns a numeric vector of length N+1: (x_0, x_1, ..., x_N)
# -------------------------------------------------------------------
metropolis_rw <- function(N = 10000, s = 1, x0 = 0) {

  x <- numeric(N + 1)   # pre-allocate N+1 slots
  x[1] <- x0            # R is 1-indexed, so x[1] = x_0

  for (i in seq(2, N + 1)) {
    proposal <- rnorm(1, mean = x[i - 1], sd = s)

    # Log-acceptance ratio
    log_r <- log_target(proposal) - log_target(x[i - 1])

    # Accept or reject
    if (log(runif(1)) < log_r) {
      x[i] <- proposal
    } else {
      x[i] <- x[i - 1]
    }
  }

  return(x)   # length N+1
}
```

## Running the Chain

We run the sampler with $N = 10000$, $s = 1$, $x_0 = 0$.

The returned vector has $N + 1 = \mathbf{10001}$ elements.  
We use $(x_1, \ldots, x_N)$ — **excluding** the initial seed $x_0$ — for
all analysis below.

```{r run_chain}
chain   <- metropolis_rw(N = 10000, s = 1, x0 = 0)  # length 10 001
samples <- chain[-1]                                  # x_1 ... x_N  (10 000 values)

cat("Full chain length  (x_0 to x_N):", length(chain),   "[should be 10 001]\n")
cat("Samples used       (x_1 to x_N):", length(samples), "[should be 10 000]\n")
```

## Visualisation: Histogram + KDE + True Laplace PDF

All three elements are overlaid in a **single figure** so the quality of the
MCMC approximation can be judged at a glance.

| Layer | Colour | Role |
|-------|--------|------|
| Histogram | light blue | Non-parametric density estimate from the samples |
| KDE | red | Smooth kernel density estimate |
| True PDF | black dashed | Theoretical Laplace(0,1) — the sampling target |

```{r plot_all_three}
library(ggplot2)

# True Laplace density on a fine grid
x_grid  <- seq(-10, 10, length.out = 800)
true_pdf <- 0.5 * exp(-abs(x_grid))
df_true  <- data.frame(x = x_grid, density = true_pdf)

# Data frame of MCMC samples
df_samples <- data.frame(x = samples)

ggplot(df_samples, aes(x = x)) +

  # 1. Histogram (density scale)
  geom_histogram(aes(y = after_stat(density)),
                 bins = 60, fill = "lightblue", colour = "white",
                 alpha = 0.7) +

  # 2. KDE
  geom_density(colour = "red", linewidth = 1, aes(linetype = "KDE")) +

  # 3. True Laplace PDF
  geom_line(data = df_true,
            aes(x = x, y = density, linetype = "True Laplace PDF"),
            colour = "black", linewidth = 1) +

  # Labels and theme
  scale_linetype_manual(name   = "Legend",
                        values = c("KDE" = "solid",
                                   "True Laplace PDF" = "dashed")) +
  labs(
    title    = "Random Walk Metropolis — Laplace(0, 1) target  (N = 10 000, s = 1, x\u2080 = 0)",
    subtitle = "Histogram of samples (blue), KDE (red), True PDF (black dashed)",
    x        = "x",
    y        = "Density"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")
```

## Monte Carlo Estimates of Mean and Standard Deviation

The sample mean and sample standard deviation are **Monte Carlo estimates**
of the true moments.

```{r stats}
sample_mean <- mean(samples)
sample_sd   <- sd(samples)

cat(sprintf("Sample mean               : %.6f\n",  sample_mean))
cat(sprintf("Sample standard deviation : %.6f\n",  sample_sd))
cat(sprintf("(True mean = 0,  true sd  = %.6f)\n", sqrt(2)))
```

The estimates are close to the true values, confirming that the sampler
has successfully explored the Laplace distribution.

---

# Part 1(b): Gelman–Rubin $\hat{R}$ Diagnostic

## Background

A single MCMC chain is not enough to confirm convergence — it could be
stuck in one region of the distribution.  The **Gelman–Rubin** diagnostic
compares **multiple chains** started from different initial values.

Given $J$ chains $\{x_i^{(j)}\}_{i=1}^{N}$, $j = 1, \ldots, J$:

**Within-chain statistics:**
$$M_j = \frac{1}{N}\sum_{i=1}^{N} x_i^{(j)}, \qquad
  V_j = \frac{1}{N}\sum_{i=1}^{N}\bigl(x_i^{(j)} - M_j\bigr)^2$$

**Between-chain statistics:**
$$W = \frac{1}{J}\sum_{j=1}^{J} V_j, \qquad
  M   = \frac{1}{J}\sum_{j=1}^{J} M_j, \qquad
  B   = \frac{1}{J}\sum_{j=1}^{J}(M_j - M)^2$$

**Diagnostic:**
$$\hat{R} = \sqrt{\frac{B + W}{W}}$$

- $\hat{R} \approx 1$ → chains agree → **converged**
- $\hat{R} > 1.05$ → chains disagree → **not converged**

## Implementation

```{r rhat_functions}
# -------------------------------------------------------------------
# compute_Rhat()
#   chains : a list of numeric vectors, each of length N+1
#            (the full chain including x_0)
# Returns the scalar R-hat value.
# -------------------------------------------------------------------
compute_Rhat <- function(chains) {

  J   <- length(chains)

  # Use x_1 ... x_N only (drop x_0 = first element)
  spc <- lapply(chains, function(ch) ch[-1])

  Mj <- sapply(spc, mean)              # within-chain means
  Vj <- sapply(spc, function(s)        # within-chain variances  (1/N, not 1/(N-1))
    mean((s - mean(s))^2))

  W    <- mean(Vj)                     # overall within-chain variance
  M    <- mean(Mj)                     # overall mean
  B    <- mean((Mj - M)^2)            # between-chain variance

  Rhat <- sqrt((B + W) / W)
  return(Rhat)
}

# -------------------------------------------------------------------
# run_chains()
#   Runs J chains from dispersed starting points and returns R-hat.
# -------------------------------------------------------------------
run_chains <- function(N = 2000, s = 0.001, J = 4) {
  starting_points <- c(-5, 0, 5, 10)
  chains <- lapply(seq_len(J),
                   function(j) metropolis_rw(N = N, s = s,
                                             x0 = starting_points[j]))
  compute_Rhat(chains)
}
```

## $\hat{R}$ for N = 2000, s = 0.001, J = 4

With $s = 0.001$ the Gaussian proposal is extremely narrow — almost every
proposal is accepted, but the chain barely moves from its starting value.
Because the four chains begin at $-5, 0, 5, 10$ and cannot mix in only
2000 steps, we expect $\hat{R} \gg 1$.

```{r rhat_specific}
set.seed(0)
rhat_s001 <- run_chains(N = 2000, s = 0.001, J = 4)
cat(sprintf("R-hat (N=2000, s=0.001, J=4) = %.4f\n", rhat_s001))
```

The large value confirms that the algorithm has **not** converged at this
choice of $s$.

## $\hat{R}$ over a Grid of 1000 s Values

Keeping $N = 2000$ and $J = 4$ fixed, we evaluate $\hat{R}$ at **1000**
equally spaced values of $s$ in $[0.001,\, 1]$.


## Plot: $\hat{R}$ vs Proposal Standard Deviation $s$

```{r rhat_grid_and_plot}
set.seed(1)
s_values    <- seq(0.001, 1, length.out = 1000)
Rhat_values <- sapply(s_values, function(s)
                 run_chains(N = 2000, s = s, J = 4))

cat(sprintf("Computed %d R-hat values.\n", length(Rhat_values)))

df_rhat <- data.frame(s = s_values, Rhat = Rhat_values)

ggplot(df_rhat, aes(x = s, y = Rhat)) +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  geom_hline(aes(yintercept = 1.05, linetype = "Threshold R-hat = 1.05"),
             colour = "red", linewidth = 1) +
  scale_linetype_manual(name = "", values = "dashed") +
  labs(
    title    = expression("Gelman\u2013Rubin " * hat(R) *
                          " vs Proposal Standard Deviation " * italic(s)),
    subtitle = "N = 2000, J = 4 chains started at x\u2080 \u2208 {-5, 0, 5, 10}",
    x        = expression("Proposal Std Dev (" * italic(s) * ")"),
    y        = expression(hat(R))
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")
```

**Interpretation:**

- For **small $s$** (near 0.001) the chains are stuck near their starting
  values and $\hat{R}$ is very large — indicating poor convergence.
- As $s$ grows the chains mix more effectively and $\hat{R}$ drops toward
  the convergence threshold of 1.05, then continues to decline toward 1.

---

# Session Information

```{r session_info}
sessionInfo()
```
