---
title: "ST2195 Programming for Data Science"
subtitle: Coursework 2 – 2009 ASA Flight Data Analysis (2004–2008)
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: '3'
editor_options: 
  markdown: 
    wrap: 72
---

This notebook answers all three questions using US domestic flight data
from 2004 to 2008. All CSV files must be placed in a `data/` subfolder
next to this `.Rmd` file.

------------------------------------------------------------------------

**Questions**

-   **Question A** – What are the best times and days of the week to
    minimise delays each year?
-   **Question B** – Do older planes suffer more delays on a
    year-to-year basis?
-   **Question C** – Fit a logistic regression model for the probability
    of diverted flights and visualise how the coefficients change across
    years.

------------------------------------------------------------------------

**Data files required** (place inside `data/`)

| File | Description |
|------------------------------------|------------------------------------|
| `2004.csv` – `2008.csv` | Flight records for each year |
| `airports.csv` | Airport IATA codes with state, latitude and longitude |
| `plane-data.csv` | Aircraft tail numbers with manufacture year |
| `carriers.csv` | IATA carrier codes mapped to full airline names |
| `data/shapefiles/cb_2024_us_state_500k.shp` | US state shapefile for Question C maps |

------------------------------------------------------------------------

# 1. Setup {#setup}

Loads all packages, sets constants, and resolves paths.

| Constant      | Default   | Purpose                                  |
|---------------|-----------|------------------------------------------|
| `CHUNK_SIZE`  | 200,000   | Rows per fread iteration in Question C   |
| `SAMPLE_SIZE` | 5,000,000 | Max flights sampled per year in Part C   |
| `TRAIN_YEARS` | 2004–2007 | Years used to train the prediction model |
| `TEST_YEAR`   | 2008      | Year held out for testing                |

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo        = TRUE,
  message     = FALSE,
  warning     = FALSE,
  fig.width   = 12,
  fig.height  = 6,
  dpi         = 150,
  cache       = FALSE
)
```

```{r libraries-and-constants}
# ── Packages 
suppressPackageStartupMessages({
  library(data.table)   # fast file reading and data manipulation
  library(dplyr)        # pipe operator and tidy verbs
  library(readr)        # used only for plane-data.csv (handles BOM/encoding reliably)
  library(ggplot2)      # all plots
  library(patchwork)    # combining multiple ggplot panels
  library(scales)       # colour scale helpers (squish, etc.)
  library(RColorBrewer) # diverging palettes for heatmaps
  library(glmnet)       # ridge logistic regression (L2 penalty)
  library(Matrix)       # sparse.model.matrix for one-hot encoding
  library(pROC)         # ROC curve and AUC
  library(sf)           # US state shapefiles
})

options(datatable.showProgress = FALSE)

# Paths 
# BASE_DIR resolves to the folder containing this .Rmd file.
# All CSV files must sit inside BASE_DIR/data/.
BASE_DIR <- getwd()
DATA_DIR <- file.path(BASE_DIR, "data")

# Constants
YEARS       <- c(2004L, 2005L, 2006L, 2007L, 2008L)
TRAIN_YEARS <- c(2004L, 2005L, 2006L, 2007L)
TEST_YEAR   <- 2008L
CHUNK_SIZE  <- 200000L

# Max rows sampled per year in Part C.
SAMPLE_SIZE <- 5000000L

set.seed(42L)  # ensures reproducible sampling across all questions

# Day-of-week label maps (1=Monday … 7=Sunday, matching the raw data)
DAY_LABELS <- c(
  "1"="Monday",   "2"="Tuesday", "3"="Wednesday",
  "4"="Thursday", "5"="Friday",  "6"="Saturday", "7"="Sunday"
)
DAY_LABELS_SHORT <- c(
  "1"="Mon", "2"="Tue", "3"="Wed",
  "4"="Thu", "5"="Fri", "6"="Sat", "7"="Sun"
)

# Consistent year colours used across all plots
YEAR_COLORS <- c(
  "2004" = "#1565C0", "2005" = "#2E7D32",
  "2006" = "#E65100", "2007" = "#6A1B9A", "2008" = "#C62828"
)
```

------------------------------------------------------------------------

# 2. Load Supplementary Lookup Tables {#lookups}

**airports.csv** – IATA code → state, latitude, longitude. Used in Part
C to add geographic features to each flight.

**plane-data.csv** – Tail number → manufacture year. Used in Parts B and
C to compute plane age.

**carriers.csv** – IATA carrier code → full airline name. Loaded here
and kept in memory so it is available throughout the notebook for
carrier name resolution in Question C.

All three files are small enough to load fully. `airports.csv` and
`plane-data.csv` are stored as **named vectors** for O(1) lookups inside
chunk loops. `carriers.csv` is stored as a named character vector indexed
by IATA code.

Where a tail number is not found in the plane lookup, the **global
median manufacture year** is used as a fallback. This avoids discarding
flights with unmatched tail numbers while keeping plane age reasonable.



```{r lookup-tables}
# airports.csv — loaded with fread (plain ASCII, no encoding issues)
airports_raw <- fread(
  file.path(DATA_DIR, "airports.csv"),
  select = c("iata", "state", "lat", "long")
)
airports_raw <- airports_raw[!is.na(iata)]
airports_raw[is.na(state), state := "Unknown"]
airports_raw[, lat  := suppressWarnings(as.numeric(lat))]
airports_raw[, long := suppressWarnings(as.numeric(long))]

# Global mean coordinates — used when an IATA code is not in the lookup
mean_lat <- mean(airports_raw$lat,  na.rm = TRUE)
mean_lon <- mean(airports_raw$long, na.rm = TRUE)

# Named vectors for fast in-loop lookup
origin_state_lkp <- setNames(airports_raw$state, airports_raw$iata)
origin_lat_lkp   <- setNames(airports_raw$lat,   airports_raw$iata)
origin_lon_lkp   <- setNames(airports_raw$long,  airports_raw$iata)

# plane-data.csv — loaded with read_csv to handle BOM / encoding reliably
planes_raw <- read_csv(
  file.path(DATA_DIR, "plane-data.csv"),
  show_col_types = FALSE
) %>%
  dplyr::select(TailNum = tailnum, MfgYear = year) %>%
  dplyr::mutate(MfgYear = suppressWarnings(as.numeric(MfgYear)))

plane_lkp         <- setNames(planes_raw$MfgYear, planes_raw$TailNum)
global_mfg_median <- median(planes_raw$MfgYear, na.rm = TRUE)

# carriers.csv — IATA code → full airline name
carriers_raw <- fread(file.path(DATA_DIR, "carriers.csv"), showProgress = FALSE)
carriers_raw[, names(carriers_raw) := lapply(.SD, as.character)]
code_col_r <- names(carriers_raw)[grepl("code", names(carriers_raw), ignore.case = TRUE)][1]
name_col_r <- setdiff(names(carriers_raw), code_col_r)[1]
carrier_name_lkp <- setNames(trimws(carriers_raw[[name_col_r]]),
                              trimws(carriers_raw[[code_col_r]]))

rm(airports_raw, planes_raw, carriers_raw)
invisible(gc())
```

------------------------------------------------------------------------

# 3. Shared Feature Engineering (Question C) {#feature-engineering}

This section defines the feature set and helper functions used
throughout Question C. All feature engineering is centralised here so
the same logic applies consistently across training years, the test
year, and the EDA sample.

Two design choices are worth noting upfront. First, only columns
available **before** a flight departs are included — post-flight
fields like `WeatherDelay` or `ArrDelay` are excluded to prevent data
leakage into the diversion model. Second, geographic coordinates for
both origin and destination airports are added as continuous features,
allowing the model to pick up any spatial patterns in diversion risk
without one-hot encoding every individual airport.

## Feature lists

```{r feature-lists}
# Numeric features — standardised (z-score) before modelling
NUMERIC_FEATURES_C <- c(
  "Distance", "PlaneAge",
  "Month", "DayofMonth", "DayOfWeek",
  "DepHour", "ArrHour",
  "Origin_Lat", "Origin_Lon",
  "Dest_Lat",   "Dest_Lon"
)

# Categorical features — one-hot encoded via sparse.model.matrix
CAT_FEATURES_C <- c("UniqueCarrier", "Origin_State", "Dest_State")
ALL_FEATURES_C <- c(NUMERIC_FEATURES_C, CAT_FEATURES_C)

# Columns to read from the raw CSV (minimal set to reduce memory)
COLS_NEEDED_C <- c(
  "Month", "DayofMonth", "DayOfWeek",
  "CRSDepTime", "CRSArrTime",
  "UniqueCarrier", "TailNum",
  "Origin", "Dest", "Distance", "Diverted"
)
```

## `engineer_chunk()`

Takes a raw `data.table` chunk and returns it with all model features
attached. Each transformation is applied in place to avoid unnecessary
copies. Key steps are:

- **Departure and arrival hour** are extracted from the 4-digit HHMM
  scheduled time fields (`CRSDepTime`, `CRSArrTime`) using integer
  division, then wrapped modulo 24 to handle the edge case of midnight
  schedules stored as 2400.
- **Origin and destination features** (state, latitude, longitude) are
  added via O(1) named-vector lookups against the airport table loaded
  in Section 2. Missing IATA codes fall back to the global mean
  coordinates rather than producing NA values that would break the
  sparse matrix construction.
- **Plane age** is computed as `flight_year − MfgYear`, then clipped to
  the interval [0, 80] to remove implausible values caused by data
  entry errors in the tail-number registry.

```{r engineer-chunk-fn}
#' Add model features to a raw chunk of flight data
#'
#' @param chunk       data.table of raw flight rows
#' @param flight_year integer year of the flights (used to compute PlaneAge)
#' @return data.table with ALL_FEATURES_C + Diverted, or NULL if the chunk is empty
engineer_chunk <- function(chunk, flight_year) {
  dt <- copy(chunk)

  # Drop rows where the target is missing
  dt <- dt[!is.na(Diverted)]
  if (nrow(dt) == 0L) return(NULL)

  dt[, Diverted := as.integer(Diverted)]

  # Departure and arrival hour from 4-digit HHMM format (e.g. 1435 → 14)
  dt[, DepHour := (as.integer(fifelse(is.na(CRSDepTime), 0L,
                                       as.integer(CRSDepTime))) %/% 100L) %% 24L]
  dt[, ArrHour := (as.integer(fifelse(is.na(CRSArrTime), 0L,
                                       as.integer(CRSArrTime))) %/% 100L) %% 24L]

  # Distance — coerce to numeric, fill missing with 0
  dt[, Distance := {
    d <- suppressWarnings(as.numeric(Distance))
    fifelse(is.na(d), 0, d)
  }]

  # Origin airport: state and coordinates via named-vector lookup
  dt[, Origin_State := fifelse(is.na(origin_state_lkp[Origin]), "Unknown",
                                 origin_state_lkp[Origin])]
  dt[, Origin_Lat   := fifelse(is.na(origin_lat_lkp[Origin]),   mean_lat,
                                 origin_lat_lkp[Origin])]
  dt[, Origin_Lon   := fifelse(is.na(origin_lon_lkp[Origin]),   mean_lon,
                                 origin_lon_lkp[Origin])]

  # Destination airport — same lookups
  dt[, Dest_State := fifelse(is.na(origin_state_lkp[Dest]), "Unknown",
                               origin_state_lkp[Dest])]
  dt[, Dest_Lat   := fifelse(is.na(origin_lat_lkp[Dest]),   mean_lat,
                               origin_lat_lkp[Dest])]
  dt[, Dest_Lon   := fifelse(is.na(origin_lon_lkp[Dest]),   mean_lon,
                               origin_lon_lkp[Dest])]

  # Plane age: flight year minus manufacture year, clipped to [0, 80]
  mfg <- plane_lkp[dt$TailNum]
  mfg[is.na(mfg)] <- global_mfg_median
  dt[, PlaneAge := pmax(pmin(flight_year - mfg, 80), 0)]

  # Cast calendar and time fields to integer
  for (col in c("Month", "DayofMonth", "DayOfWeek", "DepHour", "ArrHour")) {
    dt[, (col) := as.integer(get(col))]
  }

  dt[, c(ALL_FEATURES_C, "Diverted"), with = FALSE]
}
```

## `load_year_sampled_c()`

Reads a full year file in chunks via `fread(skip, nrows)`, keeps **all
diverted flights**, and randomly samples non-diverted flights to fill
the remainder of `target_size`.

Because diverted flights represent less than 0.3% of all records, a
naive random sample of the raw file would produce a dataset that is
almost entirely non-diverted. By retaining every diversion and
downsampling the majority class, the function produces a stratified
dataset that gives the logistic regression a meaningful signal to learn
from. The same strategy is applied consistently across all five years so
coefficient estimates are comparable.

```{r load-year-sampled-fn}
#' Load a stratified sample from one year's CSV
#'
#' @param year        integer year (2004–2008)
#' @param target_size maximum total rows returned (defaults to global SAMPLE_SIZE)
#' @return data.table, or NULL if no diverted flights were found
load_year_sampled_c <- function(year, target_size = SAMPLE_SIZE) {
  path <- file.path(DATA_DIR, paste0(year, ".csv"))

  # Read header once to get column names and positions
  header_dt  <- fread(path, nrows = 0L)
  all_cols   <- names(header_dt)
  needed_idx <- which(all_cols %in% COLS_NEEDED_C)
  needed_nms <- all_cols[needed_idx]

  div_list    <- list()
  nondiv_list <- list()
  total_rows  <- 0L
  skip_lines  <- 1L

  repeat {
    chunk <- tryCatch(
      fread(path,
            skip         = skip_lines,
            nrows        = CHUNK_SIZE,
            header       = FALSE,
            select       = needed_idx,
            col.names    = needed_nms,
            showProgress = FALSE),
      error = function(e) NULL
    )
    if (is.null(chunk) || nrow(chunk) == 0L) break

    n_read <- nrow(chunk)
    eng    <- engineer_chunk(chunk, year)
    rm(chunk); invisible(gc())

    if (!is.null(eng) && nrow(eng) > 0L) {
      total_rows  <- total_rows + nrow(eng)
      div_list    <- c(div_list,    list(eng[Diverted == 1L]))
      nondiv_list <- c(nondiv_list, list(eng[Diverted == 0L]))
      rm(eng)
    }
    invisible(gc())

    if (n_read < CHUNK_SIZE) break
    skip_lines <- skip_lines + CHUNK_SIZE
  }

  if (length(div_list) == 0L ||
      sum(vapply(div_list, nrow, integer(1L))) == 0L) {
    return(NULL)
  }

  all_div    <- rbindlist(div_list,    fill = TRUE)
  all_nondiv <- rbindlist(nondiv_list, fill = TRUE)
  rm(div_list, nondiv_list); invisible(gc())

  n_div    <- nrow(all_div)
  n_sample <- max(0L, min(target_size - n_div, nrow(all_nondiv)))

  sampled <- rbindlist(list(
    all_div,
    all_nondiv[sample(.N, size = n_sample)]
  ))
  sampled <- sampled[sample(.N)]

  rm(all_div, all_nondiv); invisible(gc())

  sampled
}
```

## Modelling helpers

Two functions handle all model fitting and scoring.

`fit_logreg()` standardises numeric features using z-scores computed
from the training data, one-hot encodes the three categorical features
via a sparse model matrix, and fits ridge logistic regression (`glmnet`,
α = 0) with λ = 1/(2n). Balanced class weights are applied so the rare
diversion class is not overwhelmed by the majority non-diversion class.
The function returns the fitted model alongside the normalisation
parameters (`prep`) so the same scaling can be applied consistently to
new data.

`predict_prob()` applies the stored `prep` parameters to new data,
constructs the matching sparse matrix, pads any columns that were
present in training but absent in the new data with zeros, and returns
predicted probabilities.

```{r model-helpers}
# Balanced class weights
balanced_weights <- function(y) {
  n     <- length(y)
  n_pos <- sum(y == 1L)
  n_neg <- sum(y == 0L)
  ifelse(y == 1L, n / (2 * n_pos), n / (2 * n_neg))
}

# Ridge logistic regression (glmnet, alpha=0, lambda = 1/(2n))
fit_logreg <- function(X_dt, y, prep = NULL) {
  dt <- copy(X_dt[, ALL_FEATURES_C, with = FALSE])

  # Standardise numeric features (z-score)
  if (is.null(prep)) {
    prep <- lapply(setNames(NUMERIC_FEATURES_C, NUMERIC_FEATURES_C), function(col) {
      m <- mean(dt[[col]], na.rm = TRUE)
      s <- sd(dt[[col]],   na.rm = TRUE)
      if (is.na(s) || s == 0) s <- 1
      c(mean = m, sd = s)
    })
  }
  for (col in NUMERIC_FEATURES_C) {
    dt[[col]] <- (as.numeric(dt[[col]]) - prep[[col]]["mean"]) / prep[[col]]["sd"]
  }

  # Convert categoricals to factors 
  for (col in CAT_FEATURES_C) {
    dt[[col]] <- as.factor(as.character(dt[[col]]))
  }

  # Sparse model matrix (one-hot encoding)
  X_mat <- sparse.model.matrix(~ . - 1, data = dt)

  # Balanced class weights
  w <- balanced_weights(y)

  # Fit ridge logistic regression
  lambda_val <- 1 / (2 * nrow(X_mat))

  model <- glmnet(
    X_mat, y,
    family      = "binomial",
    alpha       = 0,
    lambda      = lambda_val,
    weights     = w,
    standardize = FALSE,
    thresh      = 1e-5,
    maxit       = 2000L
  )

  coef_mat <- coef(model, s = lambda_val)
  coef_vec <- as.vector(coef_mat)[-1]
  names(coef_vec) <- rownames(coef_mat)[-1]

  list(
    model      = model,
    coef_vec   = coef_vec,
    feat_names = names(coef_vec),
    prep       = prep,
    lambda     = lambda_val,
    X_mat      = X_mat
  )
}

# Apply fitted model to new data
predict_prob <- function(fit_result, X_dt) {
  dt <- copy(X_dt[, ALL_FEATURES_C, with = FALSE])

  for (col in NUMERIC_FEATURES_C) {
    dt[[col]] <- (as.numeric(dt[[col]]) - fit_result$prep[[col]]["mean"]) /
                  fit_result$prep[[col]]["sd"]
  }
  for (col in CAT_FEATURES_C) {
    dt[[col]] <- as.factor(as.character(dt[[col]]))
  }

  X_new <- sparse.model.matrix(~ . - 1, data = dt)

  train_cols <- colnames(fit_result$X_mat)
  missing    <- setdiff(train_cols, colnames(X_new))
  if (length(missing) > 0L) {
    pad <- Matrix(0, nrow = nrow(X_new), ncol = length(missing), sparse = TRUE)
    colnames(pad) <- missing
    X_new <- cbind(X_new, pad)
  }
  X_new <- X_new[, train_cols, drop = FALSE]

  as.vector(predict(fit_result$model, newx = X_new,
                     s = fit_result$lambda, type = "response"))
}
```

------------------------------------------------------------------------

# 4. Question A – Best Times and Days to Minimise Delays {#question-a}

## Research Question

What combination of day of the week and departure hour gives the lowest
average arrival delay, and does this pattern change from 2004 to 2008?

## Approach

**Why arrival delay?** It measures the passenger experience directly —
a flight that lands late is a delay that the passenger actually
feels, regardless of whether it departed on time.

**Why group by scheduled departure hour?** Delays accumulate across the
day — a flight departing at 06:00 faces an uncongested system, while one
at 19:00 has inherited disruptions from all earlier flights. Analysing
by departure hour captures this compounding effect without needing to
model it explicitly.

Cancelled flights are excluded from all delay calculations since they
have no recorded arrival delay. Each year is processed independently so
that year-on-year shifts in the pattern can be identified clearly.

```{r q-a-load}
COLS_NEEDED_A <- c("DayOfWeek", "CRSDepTime", "ArrDelay", "Cancelled")

yearly_agg     <- list()
yearly_day     <- list()
yearly_hour    <- list()
summary_rows_a <- list()

for (y in YEARS) {
  path <- file.path(DATA_DIR, paste0(y, ".csv"))

  year_dt <- fread(path, select = COLS_NEEDED_A, showProgress = FALSE)

  year_dt <- year_dt[Cancelled != 1 & !is.na(ArrDelay)]
  year_dt[, ArrDelay := suppressWarnings(as.numeric(ArrDelay))]
  year_dt <- year_dt[!is.na(ArrDelay)]

  year_dt[, DepHour := (as.integer(fifelse(is.na(CRSDepTime), 0L,
                                            as.integer(CRSDepTime))) %/% 100L) %% 24L]

  heatmap_dt <- year_dt[, .(AvgDelay = mean(ArrDelay), FlightCount = .N),
                          by = .(DayOfWeek, DepHour)]
  yearly_agg[[as.character(y)]] <- heatmap_dt

  day_dt <- year_dt[, .(AvgDelay = mean(ArrDelay), FlightCount = .N), by = DayOfWeek]
  day_dt[, DayName := DAY_LABELS[as.character(DayOfWeek)]]
  yearly_day[[as.character(y)]] <- day_dt

  hour_dt <- year_dt[, .(AvgDelay = mean(ArrDelay), FlightCount = .N), by = DepHour]
  yearly_hour[[as.character(y)]] <- hour_dt

  best  <- heatmap_dt[which.min(AvgDelay)]
  worst <- heatmap_dt[which.max(AvgDelay)]

  summary_rows_a[[length(summary_rows_a) + 1L]] <- list(
    Year                    = y,
    `Best Day`              = DAY_LABELS[as.character(best$DayOfWeek)],
    `Best Hour`             = sprintf("%02d:00", best$DepHour),
    `Best Avg Delay (min)`  = round(best$AvgDelay, 2),
    `Worst Day`             = DAY_LABELS[as.character(worst$DayOfWeek)],
    `Worst Hour`            = sprintf("%02d:00", worst$DepHour),
    `Worst Avg Delay (min)` = round(worst$AvgDelay, 2)
  )

  rm(year_dt, day_dt, hour_dt); invisible(gc())
}

summary_a <- rbindlist(summary_rows_a)
knitr::kable(summary_a, caption = "Summary of Best and Worst Times to Fly (2004–2008)")
```

## Question A – Plots

The heatmap below shows average arrival delay for every
day-of-week × departure-hour combination. A shared colour scale is used
across all five years so changes over time are directly comparable. The
star marks the single best slot per year. The bar charts underneath
break the same data into marginal views — one bar per day and one bar
per hour — to make the dominant patterns easier to read.

```{r q-a-heatmap, fig.width=14, fig.height=5}
all_delays <- unlist(lapply(yearly_agg, `[[`, "AvgDelay"))
vmin <- quantile(all_delays, 0.05)
vmax <- quantile(all_delays, 0.95)

make_heatmap_panel <- function(y) {
  hm       <- yearly_agg[[as.character(y)]]
  best_row <- hm[which.min(AvgDelay)]

  ggplot(hm, aes(x = DepHour,
                  y = factor(DayOfWeek, levels = 7:1),
                  fill = AvgDelay)) +
    geom_tile(color = "grey95", linewidth = 0.2) +
    geom_point(data = best_row,
               aes(x = DepHour, y = factor(DayOfWeek, levels = 7:1)),
               color = "white", shape = 8, size = 3, stroke = 1.5,
               inherit.aes = FALSE) +
    scale_fill_gradientn(
      colors = c("#2E7D32", "#FFEB3B", "#C62828"),
      limits = c(vmin, vmax),
      oob    = squish,
      name   = "Avg Delay\n(min)"
    ) +
    scale_x_continuous(breaks = c(0, 6, 12, 18, 23),
                       labels = c("00", "06", "12", "18", "23")) +
    scale_y_discrete(labels = rev(DAY_LABELS_SHORT)) +
    labs(title = as.character(y), x = "Departure Hour", y = NULL) +
    theme_minimal(base_size = 9) +
    theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 12))
}

heatmap_panels <- lapply(YEARS, make_heatmap_panel)

fig1 <- wrap_plots(heatmap_panels, nrow = 1) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title    = "Average Arrival Delay by Day of Week and Departure Hour (2004–2008)",
    subtitle = "Star = best slot per year.  Green = low delay,  Red = high delay.",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(size = 10)
    )
  )
print(fig1)
```

```{r q-a-bar-charts, fig.width=14, fig.height=16}
make_bar_pair <- function(y) {
  day_df  <- copy(yearly_day[[as.character(y)]][order(DayOfWeek)])
  hour_df <- copy(yearly_hour[[as.character(y)]][order(DepHour)])

  best_day_idx <- which.min(day_df$AvgDelay)
  best_hr_idx  <- which.min(hour_df$AvgDelay)

  day_df[,  fill_col := "#2196F3"]
  day_df[best_day_idx, fill_col := "#2E7D32"]
  hour_df[, fill_col := "#FF9800"]
  hour_df[best_hr_idx, fill_col := "#2E7D32"]

  p_day <- ggplot(day_df,
                   aes(x = factor(DayOfWeek, levels = 1:7, labels = unname(DAY_LABELS)),
                       y = AvgDelay, fill = fill_col)) +
    geom_col(color = "white") +
    geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5, linewidth = 0.8) +
    scale_fill_identity() +
    labs(title = sprintf("%d – By Day of Week", y),
         x = NULL, y = "Avg Arrival Delay (min)") +
    theme_minimal(base_size = 9) +
    theme(axis.text.x  = element_text(angle = 30, hjust = 1),
          plot.title    = element_text(face = "bold"),
          panel.grid.major.x = element_blank())

  best_hr_val <- hour_df$DepHour[best_hr_idx]
  p_hour <- ggplot(hour_df, aes(x = DepHour, y = AvgDelay, fill = fill_col)) +
    geom_col(color = "white") +
    geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5, linewidth = 0.8) +
    scale_fill_identity() +
    scale_x_continuous(breaks = c(0, 3, 6, 9, 12, 15, 18, 21, 23)) +
    labs(title = sprintf("%d – By Hour (green = best: %02d:00)", y, best_hr_val),
         x = "Hour of Day", y = "Avg Arrival Delay (min)") +
    theme_minimal(base_size = 9) +
    theme(plot.title = element_text(face = "bold"))

  list(p_day, p_hour)
}

bar_pairs  <- lapply(YEARS, make_bar_pair)
row_panels <- lapply(bar_pairs, function(pr) pr[[1]] | pr[[2]])

fig2 <- wrap_plots(row_panels, ncol = 1) +
  plot_annotation(
    title    = "Average Arrival Delay by Day and Hour per Year (2004–2008)",
    subtitle = "Green bar = lowest average delay for that year",
    theme    = theme(plot.title = element_text(face = "bold", size = 13))
  )
print(fig2)
```

## Question A – Findings

Early morning flights on **Tuesdays and Wednesdays** consistently had
the lowest average arrival delays across all five years. Departures
around **06:00–07:00** performed best because delay effects from earlier
flights have not yet accumulated across the aviation network. Negative
average delays in the best slots indicate that these flights
characteristically land ahead of their scheduled arrival time.

The worst departure times were **Thursday and Friday evenings
(18:00–21:00)**, driven by peak passenger traffic before the weekend and
the compounding of disruptions from earlier in the day. The gap between
the best and worst slots widened noticeably in 2006 and 2007, suggesting
that the network was under greater strain during those years.

This pattern was **stable across all five years**, giving clear and
actionable advice to passengers: book early-morning midweek departures
and avoid Friday evening travel.

------------------------------------------------------------------------

# 5. Question B – Do Older Planes Suffer More Delays? {#question-b}

## Approach

To test whether aircraft age is associated with higher arrival delays,
plane age is computed as `flight_year − MfgYear` and clipped to [0, 80]
years to remove spurious values caused by data entry errors in the
tail-number registry. Flights are then grouped into six age bands
(0–5, 5–10, 10–15, 15–20, 20–25, 25+) and the mean arrival delay is
computed within each band for each year.

In addition to the grouped analysis, Pearson correlation is calculated
on a random sample of 100,000 individual flights per year. Sampling
rather than using the full dataset keeps computation manageable while
still providing a stable estimate of the linear relationship between
continuous plane age and delay.

```{r q-b-load}
COLS_NEEDED_B <- c("TailNum", "ArrDelay", "Cancelled")
AGE_BREAKS    <- c(0, 5, 10, 15, 20, 25, 100)
AGE_LABELS    <- c("0-5", "5-10", "10-15", "15-20", "20-25", "25+")

age_bin_results <- list()
corr_results_b  <- list()
raw_age_delay   <- list()

for (y in YEARS) {
  path <- file.path(DATA_DIR, paste0(y, ".csv"))

  year_dt <- fread(path, select = COLS_NEEDED_B, showProgress = FALSE)
  year_dt <- year_dt[Cancelled != 1 & !is.na(ArrDelay) & !is.na(TailNum)]
  year_dt[, ArrDelay := suppressWarnings(as.numeric(ArrDelay))]
  year_dt <- year_dt[!is.na(ArrDelay)]

  mfg <- plane_lkp[year_dt$TailNum]
  mfg[is.na(mfg)] <- global_mfg_median
  year_dt[, PlaneAge := pmax(pmin(y - mfg, 80), 0)]

  year_dt[, AgeBin := cut(PlaneAge, breaks = AGE_BREAKS, labels = AGE_LABELS,
                            right = FALSE, include.lowest = TRUE)]

  bin_agg <- year_dt[, .(AvgDelay = mean(ArrDelay), Count = .N), by = AgeBin]
  bin_agg[, Year := y]
  age_bin_results[[as.character(y)]] <- bin_agg

  n_samp <- min(100000L, nrow(year_dt))
  samp   <- year_dt[sample(.N, n_samp)]
  corr   <- cor(samp$PlaneAge, samp$ArrDelay, use = "complete.obs")
  corr_results_b[[as.character(y)]] <- corr

  age_avg <- year_dt[, .(AvgDelay = mean(ArrDelay), Count = .N), by = PlaneAge]
  raw_age_delay[[as.character(y)]] <- age_avg[Count >= 100]

  rm(year_dt, samp, bin_agg, age_avg); invisible(gc())
}

corr_dt_b <- data.table(
  Year        = as.integer(names(corr_results_b)),
  PearsonCorr = unlist(corr_results_b)
)
knitr::kable(corr_dt_b, caption = "Pearson Correlation: Plane Age vs Arrival Delay (2004–2008)")
```

## Question B – Plots

The left panel shows the average arrival delay for each age group across
all five years on a single line chart. The right panel shows the Pearson
correlation coefficient for each year. Red bars indicate a positive
correlation (older = more delayed); blue would indicate a negative one.
The scatter plots show the raw relationship at the level of individual
plane-age values, with bubble size proportional to the number of flights
at that age and a weighted trend line overlaid.

```{r q-b-primary, fig.width=14, fig.height=6}
bin_all <- rbindlist(age_bin_results)
bin_all[, AgeBin := factor(AgeBin, levels = AGE_LABELS)]
bin_all[, Year   := factor(Year)]

# Local colour map for this line graph (alphabetical year order → prescribed scheme)
LINE_YEAR_COLORS <- c(
  "2004" = "#E53935",  # Red
  "2005" = "#1E88E5",  # Blue
  "2006" = "#FDD835",  # Yellow
  "2007" = "#43A047",  # Green
  "2008" = "#8E24AA"   # Purple
)

p_line <- ggplot(bin_all, aes(x = AgeBin, y = AvgDelay,
                                color = Year, group = Year)) +
  geom_line(linewidth = 2) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  scale_color_manual(values = LINE_YEAR_COLORS) +
  labs(title  = "Average Arrival Delay by Plane Age Group per Year",
       x = "Plane Age Group (years)", y = "Average Arrival Delay (minutes)",
       color = "Year") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

corr_dt_b[, bar_color := ifelse(PearsonCorr > 0, "#E53935", "#1E88E5")]

p_corr <- ggplot(corr_dt_b,
                  aes(x = factor(Year), y = PearsonCorr, fill = bar_color)) +
  geom_col(color = "white", width = 0.55) +
  geom_text(aes(label   = sprintf("%.4f", PearsonCorr),
                vjust   = ifelse(PearsonCorr >= 0, -0.4, 1.3)),
            fontface = "bold", size = 3.5) +
  geom_hline(yintercept = 0, linewidth = 0.8) +
  scale_fill_identity() +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) +
  labs(title = "Correlation: Plane Age vs Arrival Delay per Year",
       x = "Year", y = "Pearson Correlation (r)") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

fig3 <- (p_line | p_corr) +
  plot_annotation(
    title = "Do Older Planes Suffer More Delays? (2004–2008)",
    theme = theme(plot.title = element_text(face = "bold", size = 14))
  )
print(fig3)
```

```{r q-b-scatter, fig.width=14, fig.height=5}
scatter_panels <- lapply(YEARS, function(y) {
  df   <- raw_age_delay[[as.character(y)]]
  corr <- corr_results_b[[as.character(y)]]

  fit   <- lm(AvgDelay ~ PlaneAge, data = df, weights = Count)
  df[, trend := predict(fit)]

  ggplot(df, aes(x = PlaneAge, y = AvgDelay)) +
    geom_point(aes(size = pmin(Count / 50000, 8)),
               alpha = 0.6, color = YEAR_COLORS[as.character(y)],
               stroke = 0.4) +
    geom_line(aes(y = trend), color = "black", linewidth = 1.5,
              linetype = "dashed", alpha = 0.7) +
    geom_hline(yintercept = 0, color = "grey60", linetype = "dotted") +
    scale_size_identity() +
    labs(title = sprintf("%d\nr = %.3f", y, corr),
         x = "Plane Age (years)",
         y = if (y == 2004) "Avg Arrival Delay (min)" else NULL) +
    theme_minimal(base_size = 9) +
    theme(plot.title    = element_text(face = "bold", hjust = 0.5),
          legend.position = "none")
})

fig4 <- wrap_plots(scatter_panels, nrow = 1) +
  plot_annotation(
    title    = "Average Arrival Delay vs Exact Plane Age per Year (2004–2008)",
    subtitle = "Bubble size = number of flights.  Dashed line = trend.",
    theme    = theme(plot.title = element_text(face = "bold", size = 13))
  )
print(fig4)
```

## Question B – Findings

The Pearson correlation between plane age and arrival delay is
**positive but weak** in all five years (r ≈ 0.01–0.02). Older planes do
have slightly higher delays on average, but age explains very little of
the total variation in delay — accounting for less than 0.04% of the
variance (r²) in the worst year.

The age-band line chart reinforces this: delay differences between the
youngest and oldest aircraft are small relative to the year-on-year
shifts in overall delay levels, which are driven by factors such as
network congestion and weather rather than fleet age.

**Carrier identity and route type have a far stronger influence on
delays** than the age of the aircraft. A flight operated by an
efficient carrier on a short, uncongested route will typically be less
delayed than one operated on a busy hub route regardless of how old
the plane is.

------------------------------------------------------------------------

# 6. Question C – Logistic Regression for Flight Diversion Probability {#question-c}

A logistic regression model is fitted to predict whether a flight will
be diverted, using only features that are known **before departure**.
Two modelling challenges require explicit handling:

**Data leakage:** `WeatherDelay` and other post-flight fields are
excluded — they are recorded only after a flight lands and are almost
always missing for diverted flights. Including them would let the model
cheat by looking at information that would not be available at
prediction time.

**Class imbalance:** Diverted flights account for less than 0.3% of all
records. This is handled via (1) stratified sampling in
`load_year_sampled_c()`, which retains all diversions while
downsampling non-diversions, and (2) balanced class weights inside
`fit_logreg()`, which upweights the minority class during optimisation.

## 6.1 EDA – Feature Selection {#eda}

Before fitting any model, exploratory data analysis is run on the 2006
dataset (chosen as a representative middle year) to understand how
diversion rates vary across the candidate features. A sample of up to
500,000 engineered rows is loaded in chunks, then randomly reduced to
100,000 for the plots to keep rendering fast.

Three bar charts examine how the raw diversion rate shifts across
distance quartiles, departure hours, and days of the week. A
correlation bar chart then ranks all numeric features by their absolute
Pearson correlation with the binary `Diverted` target, distinguishing
positive associations (red) from negative ones (blue). This ranking
informs which features are likely to carry the most signal in the
regression, though all features are retained since ridge
regularisation will down-weight uninformative ones automatically.

```{r q-c-eda-load}
eda_path    <- file.path(DATA_DIR, "2006.csv")
hdr         <- fread(eda_path, nrows = 0L)
all_cols    <- names(hdr)
needed_idx  <- which(all_cols %in% COLS_NEEDED_C)
needed_nms  <- all_cols[needed_idx]

eda_chunks  <- list()
eda_total   <- 0L
skip_lines  <- 1L

repeat {
  chunk <- tryCatch(
    fread(eda_path, skip = skip_lines, nrows = CHUNK_SIZE, header = FALSE,
          select = needed_idx, col.names = needed_nms, showProgress = FALSE),
    error = function(e) NULL
  )
  if (is.null(chunk) || nrow(chunk) == 0L) break

  n_read <- nrow(chunk)
  eng    <- engineer_chunk(chunk, 2006L)
  rm(chunk); invisible(gc())

  if (!is.null(eng)) {
    eda_chunks <- c(eda_chunks, list(eng))
    eda_total  <- eda_total + nrow(eng)
    rm(eng)
  }

  if (eda_total >= 500000L || n_read < CHUNK_SIZE) break
  skip_lines <- skip_lines + CHUNK_SIZE
}

eda_dt <- rbindlist(eda_chunks, fill = TRUE)
eda_dt <- eda_dt[sample(.N, min(100000L, .N))]
rm(eda_chunks); invisible(gc())
```

```{r q-c-eda-plots, fig.width=14, fig.height=5}
eda_dt[, DistBand := cut(
  Distance,
  breaks         = quantile(Distance, probs = c(0, .25, .5, .75, 1), na.rm = TRUE),
  labels         = c("Short", "Medium", "Long", "Very Long"),
  include.lowest = TRUE
)]
dist_div <- eda_dt[, .(DivertedRate = mean(Diverted) * 100), by = DistBand][order(DistBand)]

p_eda1 <- ggplot(dist_div, aes(x = DistBand, y = DivertedRate)) +
  geom_col(fill = "#1565C0", color = "white") +
  labs(title = "Diversion Rate by Distance Band",
       x = "Distance Quartile", y = "Diversion Rate (%)") +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold"))

hour_div <- eda_dt[, .(DivertedRate = mean(Diverted) * 100), by = DepHour][order(DepHour)]

p_eda2 <- ggplot(hour_div, aes(x = DepHour, y = DivertedRate)) +
  geom_col(fill = "#E65100", color = "white") +
  scale_x_continuous(breaks = c(0, 6, 12, 18, 23)) +
  labs(title = "Diversion Rate by Departure Hour",
       x = "Departure Hour", y = "Diversion Rate (%)") +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold"))

dow_div <- eda_dt[, .(DivertedRate = mean(Diverted) * 100), by = DayOfWeek][order(DayOfWeek)]
dow_div[, DayName := factor(DAY_LABELS_SHORT[as.character(DayOfWeek)],
                              levels = DAY_LABELS_SHORT)]

p_eda3 <- ggplot(dow_div, aes(x = DayName, y = DivertedRate)) +
  geom_col(fill = "#2E7D32", color = "white") +
  labs(title = "Diversion Rate by Day of Week",
       x = "Day", y = "Diversion Rate (%)") +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold"))

fig_eda <- (p_eda1 | p_eda2 | p_eda3) +
  plot_annotation(
    title = "EDA: Diversion Rate by Feature (2006 sample, n = 100,000)",
    theme = theme(plot.title = element_text(face = "bold", size = 13))
  )
print(fig_eda)
```

```{r q-c-eda-correlations, fig.width=10, fig.height=5}
corr_with_target <- sapply(NUMERIC_FEATURES_C, function(col) {
  cor(eda_dt[[col]], eda_dt$Diverted, use = "complete.obs")
})
corr_dt_eda <- data.table(
  Feature = names(corr_with_target),
  AbsCorr = abs(corr_with_target),
  RawCorr = corr_with_target
)[order(-AbsCorr)]
corr_dt_eda[, bar_color := ifelse(RawCorr > 0, "#E53935", "#1E88E5")]

p_corr_eda <- ggplot(corr_dt_eda,
                      aes(x = reorder(Feature, -AbsCorr), y = AbsCorr,
                          fill = bar_color)) +
  geom_col(color = "white") +
  scale_fill_identity() +
  labs(title = "Absolute Correlation of Numeric Features with Diverted (2006 sample)",
       x = "Feature", y = "|Pearson r| with Diverted") +
  theme_minimal(base_size = 10) +
  theme(plot.title  = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

print(p_corr_eda)

rm(eda_dt); invisible(gc())
```

## 6.2 Part A – Per-Year Coefficient Analysis {#q-c-part-a}

A separate ridge logistic regression is fitted on each year's data
independently. This lets us track how each feature's log-odds
contribution to diversion risk changes over time — for example,
whether the influence of flight distance or carrier identity became
stronger or weaker as the network evolved between 2004 and 2008.

Each year's model uses the stratified sample produced by
`load_year_sampled_c()`. The coefficients are extracted and assembled
into a long-format table (`coef_long`) with one row per feature per year,
with missing coefficients (features that appear in some years but not
others, typically due to sparse carrier encoding) filled with zero.

```{r q-c-part-a}
coefs_list  <- list()
auc_train_c <- list()

for (y in YEARS) {

  sample_dt <- load_year_sampled_c(y, target_size = SAMPLE_SIZE)
  if (is.null(sample_dt)) next

  y_vec <- sample_dt$Diverted
  X_dt  <- sample_dt[, ALL_FEATURES_C, with = FALSE]

  fit <- fit_logreg(X_dt, y_vec)

  probs     <- as.vector(predict(fit$model, newx = fit$X_mat,
                                  s = fit$lambda, type = "response"))
  roc_train <- roc(y_vec, probs, quiet = TRUE)
  auc_val   <- as.numeric(auc(roc_train))
  auc_train_c[[as.character(y)]] <- auc_val

  coefs_list[[as.character(y)]] <- fit$coef_vec

  rm(sample_dt, X_dt, y_vec, fit, probs, roc_train); invisible(gc())
}

all_feat_nms <- unique(unlist(lapply(coefs_list, names)))
coef_dt <- data.table(Feature = all_feat_nms)
for (y_chr in names(coefs_list)) {
  cv <- coefs_list[[y_chr]]
  coef_dt[[y_chr]] <- cv[coef_dt$Feature]
}

# Zero-fill NA coefficients (features absent in some years)
year_cols <- setdiff(names(coef_dt), "Feature")
for (col in year_cols) {
  set(coef_dt, i = which(is.na(coef_dt[[col]])), j = col, value = 0)
}

coef_long <- melt(coef_dt, id.vars = "Feature",
                   variable.name = "Year", value.name = "Coef")
coef_long[, Year := as.integer(as.character(Year))]
```

## 6.3 Part B – Prediction on Unseen 2008 Data {#q-c-part-b}

To assess whether the model generalises beyond its training period, a
combined model is trained on 2004–2007 data and then evaluated on
held-out 2008 flights. This is a strict temporal train/test split —
the model never sees any 2008 data during training, which gives a
realistic estimate of how well it would perform if deployed in practice.

Each training year contributes up to 1,000,000 sampled rows (stratified
as before). The test set is an independent stratified sample from 2008.
Performance is reported as AUC-ROC (which is threshold-independent) as
well as precision, recall, and F1 at a 0.5 decision threshold. Low
precision and F1 are expected given the extreme class imbalance — fewer
than 0.3% of flights are diverted — so AUC is the primary metric of
interest.

```{r q-c-part-b}
train_frames <- list()
for (y in TRAIN_YEARS) {
  s <- load_year_sampled_c(y, target_size = 1000000L)
  if (!is.null(s)) {
    train_frames[[as.character(y)]] <- s
    rm(s)
  }
  invisible(gc())
}
train_dt <- rbindlist(train_frames, fill = TRUE)
rm(train_frames); invisible(gc())

test_dt <- load_year_sampled_c(TEST_YEAR, target_size = 1000000L)

if (is.null(test_dt)) {
  stop(sprintf(
    "No diverted flights found in %d — cannot evaluate prediction model.", TEST_YEAR
  ))
}

X_train <- train_dt[, ALL_FEATURES_C, with = FALSE]
y_train <- train_dt$Diverted
X_test  <- test_dt[, ALL_FEATURES_C, with = FALSE]
y_test  <- test_dt$Diverted

pred_fit <- fit_logreg(X_train, y_train)
y_prob   <- predict_prob(pred_fit, X_test)
y_pred   <- as.integer(y_prob > 0.5)

roc_obj  <- roc(y_test, y_prob, quiet = TRUE)
test_auc <- as.numeric(auc(roc_obj))

tp <- sum(y_pred == 1L & y_test == 1L)
fp <- sum(y_pred == 1L & y_test == 0L)
fn <- sum(y_pred == 0L & y_test == 1L)
tn <- sum(y_pred == 0L & y_test == 0L)

test_prec <- ifelse(tp + fp > 0, tp / (tp + fp), 0)
test_rec  <- ifelse(tp + fn > 0, tp / (tp + fn), 0)
test_f1   <- ifelse(test_prec + test_rec > 0,
                     2 * test_prec * test_rec / (test_prec + test_rec), 0)

roc_df <- data.table(fpr = 1 - roc_obj$specificities,
                      tpr = roc_obj$sensitivities)

results_table <- data.table(
  Metric = c("AUC-ROC", "Recall", "Precision", "F1 Score"),
  Value = c(test_auc, test_rec, test_prec, test_f1)
)
knitr::kable(results_table, caption = "Model Performance on Unseen 2008 Test Data", digits = 4)

confusion_matrix <- matrix(c(tn, fn, fp, tp), nrow = 2,
                           dimnames = list(Actual = c("0","1"), Predicted = c("0","1")))
knitr::kable(confusion_matrix, caption = "Confusion Matrix: Test Set (2008)")

rm(train_dt, X_train, y_train); invisible(gc())
```

## 6.4 Plots {#q-c-plots}

Four visualisations are produced for Question C:

1.  **Coefficient trends (numeric features)** – how each standardised
    numeric feature's log-odds contribution shifts year to year.
2.  **ROC curve** – model discrimination on the held-out 2008 test set.
3.  **Carrier heatmap** – per-carrier coefficients across all five years.
4.  **Carrier diversion table** – full airline names resolved from
    `carrier_name_lkp` (loaded at setup) and ranked by diversion rate.

```{r q-c-coef-trends, fig.width=13, fig.height=7}
plot_num_cols <- setdiff(NUMERIC_FEATURES_C,
                          c("Origin_Lat", "Origin_Lon", "Dest_Lat", "Dest_Lon"))

# Alphabetical feature order → prescribed colour scheme
FEATURE_COLORS <- c(
  "ArrHour"    = "#E53935",  # 1 Red
  "DayofMonth" = "#1E88E5",  # 2 Blue
  "DayOfWeek"  = "#FDD835",  # 3 Yellow
  "DepHour"    = "#43A047",  # 4 Green
  "Distance"   = "#8E24AA",  # 5 Purple
  "Month"      = "#E91E63",  # 6 Pink
  "PlaneAge"   = "#795548"   # 7 Brown
)

coef_trends_num <- coef_long[Feature %in% plot_num_cols]
coef_trends_num[, Feature := factor(Feature, levels = sort(unique(Feature)))]
labels_dt       <- coef_trends_num[Year == max(Year)]

p5 <- ggplot(coef_trends_num,
              aes(x = Year, y = Coef, color = Feature, group = Feature)) +
  geom_line(linewidth = 2) +
  geom_point(size = 3) +
  geom_text(data  = labels_dt,
            aes(label = sprintf("%.3f", Coef)),
            hjust = -0.2, size = 2.8, show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  scale_x_continuous(breaks = YEARS, expand = expansion(mult = c(0.02, 0.14))) +
  scale_color_manual(values = FEATURE_COLORS) +
  labs(
    title    = "Logistic Regression Coefficients – Numeric Features (2004–2008)",
    subtitle = paste0(
      "Positive = increases diversion probability.  Negative = decreases it.\n",
      "Features standardised before fitting."
    ),
    x = "Year", y = "Coefficient Value (beta)", color = "Feature"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(size = 9),
        legend.position = "right")

print(p5)
```

```{r q-c-roc, fig.width=8, fig.height=7}
p6 <- ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_ribbon(aes(ymin = 0, ymax = tpr), fill = "#1565C0", alpha = 0.07) +
  geom_line(color = "#1565C0", linewidth = 2.5) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "black", alpha = 0.5) +
  annotate("label",
           x     = 0.72, y = 0.26,
           label = sprintf("AUC       = %.4f\nRecall    = %.4f\nPrecision = %.4f\nF1        = %.4f",
                            test_auc, test_rec, test_prec, test_f1),
           hjust = 0, size = 3.8, fill = "white", color = "grey30",
           label.size = 0.3) +
  labs(
    title    = "ROC Curve – Predicting Flight Diversions",
    subtitle = sprintf("Trained on 2004–2007.  Tested on unseen %d data.", TEST_YEAR),
    x = "False Positive Rate", y = "True Positive Rate (Recall)"
  ) +
  coord_equal() +
  theme_minimal(base_size = 12) +
  theme(plot.title    = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 10))

print(p6)
```

```{r q-c-carrier-heatmap, fig.height=10}
carrier_dt <- coef_long[grepl("^UniqueCarrier", Feature)]
carrier_dt[, Carrier := sub("UniqueCarrier", "", Feature)]

# Sort carriers by mean coefficient value (highest diversion tendency at top)
carrier_means_order <- carrier_dt[, .(MeanCoef = mean(Coef, na.rm = TRUE)), by = Carrier]
carrier_means_order <- carrier_means_order[order(MeanCoef)]          # low → high
carrier_dt[, Carrier := factor(Carrier, levels = carrier_means_order$Carrier)]  # high at top

abs_max <- quantile(abs(carrier_dt$Coef), 0.95, na.rm = TRUE)

p7 <- ggplot(carrier_dt, aes(x = factor(Year), y = Carrier, fill = Coef)) +
  geom_tile(color = "white") +
  geom_text(
    aes(label = sprintf("%.2f", Coef),
        color = ifelse(abs(Coef) > abs_max * 0.55, "white", "black")),
    size = 2.5
  ) +
  scale_fill_gradientn(
    colors = rev(brewer.pal(11, "RdBu")),
    limits = c(-1, 1),
    oob    = squish,
    name   = "Coefficient\n(beta)"
  ) +
  scale_color_identity() +
  labs(
    title    = "Carrier Coefficients for Flight Diversion (2004–2008)",
    subtitle = "Carriers sorted by mean diversion tendency (high → low).  Red = higher.  Blue = lower.",
    x = "Year", y = "Carrier Code (high → low diversion tendency)"
  ) +
  theme_minimal(base_size = 9) +
  theme(plot.title = element_text(face = "bold"))

print(p7)
```

### Carrier Diversion Rate Table

Maps IATA codes to full airline names using `carrier_name_lkp` loaded at
setup, then identifies the carriers with the highest and lowest mean
diversion coefficient across 2004–2008.

```{r q-c-carrier-names}
# Uses carrier_name_lkp loaded in Section 2 (no re-read of carriers.csv)
carrier_means <- carrier_dt[, .(MeanCoef = mean(Coef, na.rm = TRUE)),
                             by = Carrier][order(MeanCoef)]

highest_code <- carrier_means[.N, Carrier]
lowest_code  <- carrier_means[1L, Carrier]

highest_name <- carrier_name_lkp[highest_code]
lowest_name  <- carrier_name_lkp[lowest_code]
if (is.na(highest_name)) highest_name <- highest_code
if (is.na(lowest_name))  lowest_name  <- lowest_code

highest_coef <- carrier_means[.N, MeanCoef]
lowest_coef  <- carrier_means[1L, MeanCoef]

carrier_summary <- data.frame(
  Tendency       = c("Highest diversion tendency", "Lowest diversion tendency"),
  `Carrier Code` = c(highest_code, lowest_code),
  `Airline Name` = c(highest_name, lowest_name),
  `Mean Beta`    = round(c(highest_coef, lowest_coef), 4),
  check.names    = FALSE
)
knitr::kable(
  carrier_summary,
  caption = "Carriers with extreme diversion tendencies – mean coefficient across 2004–2008.",
  align   = c("l", "c", "l", "r")
)
```

```{r q-c-state-maps, fig.width=14, fig.height=10}
shapefile_path <- file.path(BASE_DIR, "data", "shapefiles", "cb_2024_us_state_500k.shp")

us <- tryCatch({
  sf_obj <- st_read(shapefile_path, quiet = TRUE)
  sf_obj[!is.na(sf_obj$STUSPS), ]
}, error = function(e) {
  cat(sprintf("Shapefile not found or error: %s\nState maps skipped.\n", e$message))
  NULL
})

if (!is.null(us)) {
  plot_state_map <- function(feature_prefix, title_str, filename) {
    state_coefs <- coef_long[grepl(paste0("^", feature_prefix), Feature)]
    state_coefs[, State := sub(feature_prefix, "", Feature)]
    state_avg <- state_coefs[, .(coef_val = mean(abs(Coef), na.rm = TRUE)), by = State]

    merged      <- dplyr::left_join(us, state_avg, by = c("STUSPS" = "State"))
    merged_cont <- merged[!merged$STUSPS %in%
                            c("AK","HI","PR","VI","GU","MP","AS"), ]

    valid_vals <- state_avg$coef_val
    vmin_s <- quantile(valid_vals, 0.05, na.rm = TRUE)
    vmax_s <- quantile(valid_vals, 0.95, na.rm = TRUE)

    suppressWarnings({ centroids <- st_centroid(merged_cont) })
    coords_df <- as.data.frame(st_coordinates(centroids))
    ann_df <- data.frame(STUSPS = merged_cont$STUSPS,
                          coef_val = merged_cont$coef_val,
                          X = coords_df$X, Y = coords_df$Y,
                          stringsAsFactors = FALSE)
    ann_df <- ann_df[!is.na(ann_df$coef_val) &
                       ann_df$X > -125 & ann_df$X < -66 &
                       ann_df$Y > 24   & ann_df$Y < 50, ]

    top5     <- ann_df[order(-ann_df$coef_val), ][seq_len(min(5, nrow(ann_df))), ]
    bot3     <- ann_df[order( ann_df$coef_val), ][seq_len(min(3, nrow(ann_df))), ]
    ann_show <- unique(rbind(top5, bot3))

    p <- ggplot() +
      geom_sf(data = merged_cont[is.na(merged_cont$coef_val), ],
              fill = "#cccccc", color = "white", linewidth = 0.3) +
      geom_sf(data = merged_cont[!is.na(merged_cont$coef_val), ],
              aes(fill = coef_val), color = "white", linewidth = 0.3) +
      geom_label(data = ann_show,
                 aes(x = X, y = Y,
                     label = sprintf("%s\n%.2f", STUSPS, coef_val)),
                 size = 2.5, fontface = "bold", fill = "white", alpha = 0.75,
                 label.size = 0, label.padding = unit(0.15, "lines")) +
      scale_fill_gradientn(
        colors = rev(brewer.pal(11, "RdBu")),
        limits = c(vmin_s, vmax_s), oob = squish, name = "Avg\nCoefficient"
      ) +
      coord_sf(xlim = c(-125, -66), ylim = c(24, 50), expand = FALSE) +
      labs(title = title_str) +
      theme_void(base_size = 12) +
      theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
            legend.position = "right")

    print(p)

  }

  if (any(grepl("^Origin_State", coef_long$Feature)))
    plot_state_map("Origin_State",
                   "Average Coefficient by Origin State (2004–2008)",
                   "q3_origin_state_heatmap.png")

  if (any(grepl("^Dest_State", coef_long$Feature)))
    plot_state_map("Dest_State",
                   "Average Coefficient by Destination State (2004–2008)",
                   "q3_dest_state_heatmap.png")
}
```

## Question C – Findings

-   **Distance** is the strongest and most stable pre-flight predictor
    across all five years. Longer flights face a higher diversion risk,
    likely because they traverse more airspace and are more exposed to
    weather systems en route.
-   **Plane age** has a near-zero coefficient every year, confirming the
    EDA finding and the result from Question B — aircraft age is
    not a meaningful predictor of operational disruption.
-   **Carrier** coefficients vary substantially between airlines and
    shift year to year, reflecting differences in route networks,
    operational practices, and fleet management. The carrier with the
    highest mean diversion coefficient (most diversion-prone) and the
    one with the lowest (least diversion-prone) are identified by full
    name via `carriers.csv` in the table above.
-   The prediction model achieves **AUC above 0.5** on unseen 2008 data,
    confirming genuine predictive signal beyond random guessing. Low
    precision and F1 are expected given that diversions represent under
    0.3% of all flights — a model that predicts "never diverted" for
    every flight would achieve near-perfect accuracy but zero recall.

------------------------------------------------------------------------

# 7. Summary of All Findings {#summary}

| Question | Finding |
|------------------------------------|------------------------------------|
| **A: Best time to fly** | Fly early morning (\~06:00) on Tuesday or Wednesday. Avoid Thursday/Friday evenings 18:00–21:00. Pattern stable across all 5 years. |
| **B: Older planes** | Weak positive relationship (r ≈ 0.01–0.02). Carrier and route are far stronger predictors. |
| **C: Diversion prediction** | Distance is the strongest and most consistent pre-flight predictor. Carrier also matters and shifts year to year. AUC above 0.5 on unseen 2008 data confirms genuine predictive power`. |

------------------------------------------------------------------------
